Potential of I/O-Aware Workflows in Climate and Weather

### 2020-02-26 paper.tex
Further comments.

This paper is structured as follows.
First, we describe the software stack involved in executing workflows in climate and weather in \Cref{sec:workflows}.
Related work in heterogeneous storage environments and solutions for workflow processing are presented in \Cref{sec:art}.
Next, the vision for including knowledge about data requirements and characteristics \lr{data requirements and characteristics == data management?} is sketched in \Cref{sec:vision} outlining the potential benefit the automatic exploitation might bring.
Our design, based on existing components in climate and weather, is described in \Cref{sec:design}.
The paper is concluded in \Cref{sec:conclusions}.

@@

  \item \textbf{Application} is executed taking the generated filenames set by the script.
\end{enumerate}

@@

\lr{Figs 3 and 4 together as an option.}

\begin{figure}[b]
    \begin{minipage}{.42\linewidth}
        \centering
        \includegraphics[width=0.42\textwidth]{pic/layers-xios}
        \subcaption{I/O path for an MPI-parallel application}\label{pic/layers-xios}
    \end{minipage}
    \begin{minipage}{.58\linewidth}
        \centering
        \includegraphics[width=1\textwidth]{pic/system}
        \subcaption{Example of an heterogeneous HPC landscape}\label{pic/system}
    \end{minipage}
    \vspace{5pt}
    \caption{I/O stack and data centre infrastructure}
    \label{fig:34}
\end{figure}

@@

There are several approaches to implement the technology for the vision proposed in this work, and changes are needed in the software components to realise it.
In the next sections, we will discuss a specific design for our transitional roadmap considering climate and weather workflows and tools scientists from this field already use in their routine research.

\lr{"a specific design" here? Next section, no?}

@@

While many optimisations are possible once an abstraction is in place, the improvements we discuss here are related to the life cycle of datasets and the placement of such datasets into specific storage according to system performance characteristics and the workflow specification.
To achieve that, the system information shall comprise of all available storage systems, the system topology, and details of each of the required components.
Simplified and complex models of the components can be included to approximate expected performance for specific I/O patterns.
It is expected that the data centre (or expert users) can create such a configuration file, e.g., by using vendor-provided information or by executing benchmarks.
With this information, a scheduler can make data placement, transformation, and migration decisions for individual datasets during their life cycle.\lr{The scheduler making data transformation? And it's migration already classified as data placement?}
This separation of concerns allows us to abstract from the workflow what is essential and what a system should optimise to ensure smart usage of the available resources.

@@

\item[Data Allocation] is the assignment of a specific area of an available storage system to particular data.
In current workflows, the user usually has a script for each task defining the filenames with a prefix that places datasets generated by the same task into a specific storage\footnote{Complicated scripts would have allowed changing the storage type depending on the cycle. Still, it is a significant burden to the user.}.
Because there is one script responsible for generating the configuration, the decision in which directory the dataset will be stored is somewhat fixed.
Such configuration is done manually and with restricted information about the system architecture.
It would be interesting to explore storage options for the datasets and, e.g., having datasets from different cycles placed at different storage systems.
For instance, in \cref{pic/workflow-s3-n}, alternating the storage location for Product\,2 into two scratch file systems is something that would be a simple job for an I/O-aware scheduler.
However, currently, that implies having at least two scripts for that task with information about the different storage placement.
%Next, we exemplify three situations in which data placement can be optimised.

\lr{Weak justification. We are better than that.}

\item[Data Migration] is the process of transferring data from one storage system to another.
Data movement involves a significant overhead, both in terms of latency and energy-efficient computing, hence needs to be considered carefully.
\cref{fig:lifecycle} introduces three possible life cycles for a specific dataset and explains how migrations can be done to improve datasets accessibility.
In \cref{fig:lifecycle1}, the dataset could be first stored on the \texttt{local} storage to avoid congestion on the \texttt{work} file system, then be migrated to \texttt{work} file system where subsequent tasks of the workflow may read it multiple times.
In the end, this dataset may be an intermediate product that can then be deleted.
In \cref{fig:lifecycle2}, the dataset is stored on the \texttt{scratch} file system immediately and accessed there.
%However, that would require subsequent tasks to be placed on the same node where data is now stored, and the last read access must happen before files on \texttt{scratch} are automatically removed.
However, the last read access must happen before files on \texttt{scratch} are automatically removed.
Alternatively, \cref{fig:lifecycle3} presents the case where the dataset is created on \texttt{work} by a task and it is copied to a \texttt{local} node.
This \texttt{local} node allows reading accesses of subsequent tasks which might be beneficial for small random accesses.
For the last two scenarios, subsequent tasks would have to be placed on the same node where previous data was stored.
%However, it would require that the subsequent tasks are placed on the same node where data is now stored.

@@

\lr{If we are really migrating, why do we have to deletes in (a) and (c)? Maybe the delete should appear at the same time as the migration in the timeline.}

\item[Data Replication] in computing involves sharing information to ensure consistency between redundant resources, such as software or hardware components, to improve reliability, fault-tolerance, or accessibility.
Data might be replicated by enabling the system to rerun parts of the workflow in case of a data loss.
In addition, the system may combine the replication of data by \textbf{transforming} the data into a different representation allowing to achieve better performance for a variety of access patterns. \lr{Talk about ESDM!}

\item[Direct-Coupling] replaces I/O by communicating data between subsequent steps of a workflow directly without storing intermediate data products on persistent storage.
As an example, in \cref{pic/workflow-s3-n}, the outcome of Task\,1 may be used directly by Task\,2.
To achieve decoupling\lr{Decoupling in the direct-coupling session? Confusing to me.} between producer and consumer, data may also be kept in memory and cached.
\end{description}

\lr{We talked about balancing the text for each item. I think there was no time for it, right?}

@@

\item Adapt the resulting workflow, individual scripts and application executions to consider the potential for data placement and migration.\lr{data placement and migration again.}

\end{enumerate}

@@

  \item \textbf{Scientist} The end-user \lr{Is the end-user the scientist?} now has to provide an additional file that covers the I/O information for each task and slightly changes have to be made to the current scripts (discussed above).\lr{Where? I'm lost again. Do we change anything in the original scripts that the user has to provide? He has to provide an additional file and then everything will be under our control, no? The scripts will change, but not by the user, by EIOS.}

@@

  \item \textbf{Job}
  A job might run on the same node than a previous job to utilise local storage.
  %\bnl{Any interesting that needs this performance is running on dozens if not hundreds of nodes}
  \lr{Only that might happen?}

@@

  \item \textbf{Application}
  The application may either use XIOS, NetCDF with ESDM support or ESDM directly to access datasets.
  Hence, in \cref{pic/layers-xios}, the HDF5 layer is replaced with ESDM.
  ESDM loads the \texttt{esdm.conf} file that contains the information about the available storage backends and their characteristics.
  ESDM extracts the long-term schedule information from the generated pseudo filenames and employs it during the I/O scheduling to optimise the storage considering data locality between tasks.\lr{Is that really the only thing we are exploring?}

  \lr{In this happening in the application level? Not before slurm? I'm confused now.}

\end{enumerate}

>>>>

This paper is structured as follows.
First, we describe the software stack involved in executing workflows in climate and weather in \Cref{sec:workflows}.
Related work in heterogeneous storage environments and solutions for workflow processing are presented in \Cref{sec:art}.
Next, the vision for including knowledge about data requirements and characteristics is sketched in \Cref{sec:vision} outlining the potential benefit the automatic exploitation might bring.
Our design, based on existing components in climate and weather, is described in \Cref{sec:design}.
The paper is concluded in \Cref{sec:conclusions}.

@@

  \item \textbf{Application} is executed taking the filenames set by the script.
\end{enumerate}

@@

\begin{figure}[b]
    \begin{minipage}{.42\linewidth}
        \centering
        \includegraphics[width=0.42\textwidth]{pic/layers-xios}
        \subcaption{I/O path for an MPI-parallel application}\label{pic/layers-xios}
    \end{minipage}
    \begin{minipage}{.58\linewidth}
        \centering
        \includegraphics[width=1\textwidth]{pic/system}
        \subcaption{Example of an heterogeneous HPC landscape}\label{pic/system}
    \end{minipage}
    \vspace{5pt}
    \caption{Typical hardware and software environment for applications}
    \label{fig:34}
\end{figure}

@@

There are several approaches to implement the technology for the vision proposed in this work, and changes are needed in the software components to realise it.
In \Cref{sec:design}, we will discuss a specific design for our transitional roadmap considering climate and weather workflows and tools scientists from this field already use in their routine research.

@@

While many optimisations are possible once an abstraction is in place, the improvements we discuss here are related to the life cycle of datasets and the placement of such datasets into specific storage according to system performance characteristics and the workflow specification.
To achieve that, the system information shall comprise of all available storage systems, the system topology, and details of each of the required components.
Simplified and complex models of the components can be included to approximate expected performance for specific I/O patterns.
It is expected that the data centre (or expert users) can create such a configuration file, e.g., by using vendor-provided information or by executing benchmarks.
With this information, a scheduler can make the initial data placement, transformation, and migration decisions for individual datasets during their life cycle.
This separation of concerns allows us to abstract from the workflow what is essential and what a system should optimise to ensure smart usage of the available resources.

@@

\item[Data Allocation] is the assignment of a specific area of an available storage system to particular data.
In current workflows, the user usually has a script for each task defining the filenames with a prefix that places datasets generated by the same task into a specific storage\footnote{Complicated scripts would have allowed changing the storage type depending on the cycle. Still, it is a significant burden to the user.}.
Because there is one script responsible for generating the configuration, the decision in which directory the dataset will be stored is somewhat fixed.
Such configuration is done manually and with restricted information about the system architecture.
It would be interesting to explore storage options for the datasets and, e.g., having datasets from different cycles placed at different storage systems.
For instance, in \cref{pic/workflow-s3-n}, alternating the storage location for Product\,2 into two scratch file systems is something that would be a simple job for an I/O-aware scheduler.
However, currently, that implies providing scripts for that task and all tasks depending on the data with information about the different storage placement.
%Next, we exemplify three situations in which data placement can be optimised.

\item[Data Migration] is the process of transferring data from one storage system to another; typically this involves to delete the data but this decision can be delayed to provide read access on multiple storage systems.
Data movement involves a significant overhead as data must be read on one storage and written to another, both in terms of latency and energy-efficient computing, hence needs to be considered carefully.
\cref{fig:lifecycle} introduces three possible life cycles for a specific dataset and explains how migrations can be done to improve datasets accessibility.
In \cref{fig:lifecycle1}, the dataset could be first stored on the \texttt{local} storage to avoid congestion on the \texttt{work} file system, then be migrated to \texttt{work} file system where subsequent tasks of the workflow may read it multiple times.
In the end, this dataset may be an intermediate product that can then be deleted.
In \cref{fig:lifecycle2}, the dataset is stored on the \texttt{scratch} file system immediately and accessed there.
%However, that would require subsequent tasks to be placed on the same node where data is now stored, and the last read access must happen before files on \texttt{scratch} are automatically removed.
However, the last read access must happen before files on \texttt{scratch} are automatically removed.
Alternatively, \cref{fig:lifecycle3} presents the case where the dataset is created on \texttt{work} by a task and it is copied to a \texttt{local} node.
This \texttt{local} node allows reading accesses of subsequent tasks which might be beneficial for small random accesses.
For the last two scenarios, subsequent tasks would have to be placed on the same node where previous data was stored.
%However, it would require that the subsequent tasks are placed on the same node where data is now stored.

@@

\item[Data Replication] in computing involves sharing information to ensure consistency between redundant resources, such as software or hardware components, to improve reliability, fault-tolerance, or accessibility.
Data might be replicated by enabling the system to rerun parts of the workflow in case of a data loss.
In addition, the system may combine the replication of data by \textbf{transforming} the data into a different representation allowing to achieve better performance for a variety of access patterns.

\item[Direct-Coupling] replaces I/O by communicating data between subsequent steps of a workflow directly without storing intermediate data products on persistent storage.
As an example, in \cref{pic/workflow-s3-n}, the outcome of Task\,1 may be used directly by Task\,2.
To achieve some level of independence between producer and consumer, data may also be kept in memory and cached.
\end{description}

@@

\item Adapt the resulting workflow, individual scripts and application executions to consider the potential for data placement.
\end{enumerate}

@@

  \item \textbf{Scientist} The user now has to provide an additional file that covers the I/O information for each task and slightly changes have to be made to the current scripts.

@@

  \item \textbf{Job}
  A job might run on the same node than a previous job to utilise local storage.
  %\bnl{Any interesting that needs this performance is running on dozens if not hundreds of nodes}

@@

  \item \textbf{Application}
  The application may either use XIOS, NetCDF with ESDM support or ESDM directly to access datasets.
  Hence, in \cref{pic/layers-xios}, the HDF5 layer is replaced with ESDM.
  ESDM loads the \texttt{esdm.conf} file that contains the information about the available storage backends and their characteristics.
  ESDM extracts the long-term schedule information from the generated pseudo filenames and employs it during the I/O scheduling to optimise the storage considering data locality between tasks.
  Basically, ESDM can now change the priorities for data placement of the different storage locations that would normally be encoded in its configuration file.
\end{enumerate}
